Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-11-11 Dail_Mad_ext.tex
abstract edits

%\doublespacing
\author{Jeffrey A. Hostetler\\ %\footnote{Contact:
%     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
        \small Migratory Bird Center \\
        \small Smithsonian Conservation Biology Institute \\
        \small National Zoological Park, MRC 5503 \\
        \small Washington, DC 20013-7012\\
   \and Richard B. Chandler \footnote{Contact:
     \url{rbchan@uga.edu}, +001-706-542-5815} \\
        \small USGS Patuxent Wildlife Research Center \\
        \small Laurel, MD 20708}
\maketitle

\abstract{
\noindent
\begin{enumerate}
\item Models of population dynamics play a central role in theoretical and
applied ecology, where they are used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Failure to
account for observation error in such models can result in 
bias. Therefore, ecologists have increasingly relied on state-space
models to directly model both the observation error and the ecological
process of interest.
\item Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) we have no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect detection.
\item We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods. We provide code to fit these models using the \textbf{R} package
\texttt{unmarked} and \textbf{JAGS} is also provided.
\item A simulation study indicated that bias was negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlighted how these methods
can be readily applied to existing data, but it also suggested 
that precision would be low when we lack direct information about detection
probability (such as is collected using distance sampling or
replicated counts).
\end{enumerate}
}

@@

\textbf{Key words}: abundance, Dail and Madsen model, density-dependence,
detectability, Gompertz-logistic model, immigration, open population point 
count models, random observer effects, Ricker-logistic model, zero-inflated

>>>>

%\doublespacing
\author{Jeffrey A. Hostetler\\ %\footnote{Contact:
%     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
        \small Migratory Bird Center \\
        \small Smithsonian Conservation Biology Institute \\
        \small National Zoological Park, MRC 5503 \\
        \small Washington, DC 20013-7012\\
   \and Richard B. Chandler \footnote{Current address: Warnell School of
     Forestry and Natural Resources, University of Georgia,
     \url{rchandler@warnell.uga.edu}, +001-706-542-5815} \\
        \small USGS Patuxent Wildlife Research Center \\
        \small Laurel, MD 20708
}
\maketitle

\abstract{
\noindent
\begin{enumerate}
\item Models of population dynamics %play a central role in theoretical and
%applied ecology, where they 
are frequently used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Fitting models
of population dynamics to field data is challenging because most data
sets are characterized by missing data and observation error, which
can inflate estimates of process variation if ignored.  
%Failure to account for observation error in such models can result in 
%bias. Therefore, ecologists have increasingly relied on 
Recently, state-space models have been developed to deal with these problems
%these probdirectly model 
by directly modeling both the observation error and the ecological
process of interest.
\item Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) they assert an implausible model for
%we have no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect detection.
\item We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code.%. We provide code to fit these models using the \textbf{R} package
%\texttt{unmarked} and \textbf{JAGS} is also provided.
\item Results of a simulation study suggest that bias is %was 
negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlights how these methods
can be readily applied to existing data, but it also suggests 
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is lacking.
\end{enumerate}
}

@@

\textbf{Key words}: abundance, Dail and Madsen model, density-dependence,
detectability, Gompertz-logistic model, immigration, open population point 
count models, population dynamics, random observer effects, Ricker-logistic model, zero-inflated
